---
title: "IntegrateATAC"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
[Eukaryotic chromatin is tightly packaged into nucleosomes, and the positioning of nucleosomes can regulate gene expression by changing the in vivo availability of binding sites to transcription factors (TFs)](https://epigeneticsandchromatin.biomedcentral.com/articles/10.1186/1756-8935-7-33). Assay of Transposase Accessible Chromatin sequencing (ATAC-seq) is a technique used to assess genome-wide chromatin accessibility. Thus, combining ATAC-seq with RNA-seq can be used to understanding the role of chromatin structure in regulating gene expression.

<hr />

## Example data
The data used here contains RNA-seq and ChIP-seq datasets from [RUNX represses Pmp22 to drive neurofibromagenesis](https://www.science.org/doi/10.1126/sciadv.aau8389): 

* RNA-seq: two genotypes and three samples per genotype, the raw data are stored in [GSE122774](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122774)
* ATAC-seq: two genotypes and one sample per genotype, the raw data are stored in [GSE122776](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122776)

## Differential Expression Analysis (RNA-seq)
The RNA-seq data used here are same as used in "IntegrateChIP" vignette.
```{r debatac_degs}
# library
suppressWarnings(suppressMessages(library(DESeq2)))
suppressWarnings(suppressMessages(library(DEbPeak)))

# prepare count matrix and metadata
debatac.count.file <- system.file("extdata", "debchip_count.txt", package = "DEbPeak")
debatac.meta.file <- system.file("extdata", "debchip_meta.txt", package = "DEbPeak")
debatac.count.matrix <- read.table(file = debatac.count.file, header = T, sep = "\t")
debatac.meta.info <- read.table(file = debatac.meta.file, header = T)
# create DESeqDataSet object
debatac.dds <- DESeq2::DESeqDataSetFromMatrix(countData = debatac.count.matrix, 
                                              colData = debatac.meta.info, 
                                              design = ~condition)

# set control level
debatac.dds$condition <- relevel(debatac.dds$condition, ref = "NF")
# conduct differential expressed genes analysis
debatac.dds <- DESeq(debatac.dds)
# extract results
debatac.dds.results <- results(debatac.dds,contrast=c("condition",'RX','NF'))
debatac.dds.results.ordered <- debatac.dds.results[order(debatac.dds.results$log2FoldChange,decreasing = TRUE),]
head(debatac.dds.results.ordered)
```

<hr />

## Process ATAC-seq data
### Consensus peaks
```{r debatac_atac_consensus_peak}
# get consensus peak
atac.peak.file <- system.file("extdata", "debatac_peaks.bed", package = "DEbPeak")
atac.peak.df <- GetConsensusPeak(peak.file = atac.peak.file)
head(atac.peak.df)
```

<hr />

### Peak profile
Check the profle of consensus peaks:
```{r debatac_atac_peak_profile, fig.height = 6, fig.width = 10, fig.align = "center", warning=FALSE}
# peak profile plot
atac.peak.profile <- PeakProfile(atac.peak.df, species = "Mouse", by = "gene", region.type = "body", nbin = 800)
atac.peak.profile$profile.plot
```

<hr />

### Peak annotation
In this step, we will perform peak annotation with [ChIPseeker](https://bioconductor.org/packages/release/bioc/vignettes/ChIPseeker/inst/doc/ChIPseeker.html). This will include annotating peaks with its nearby genes, assigning genomic region of the peaks, et al.
```{r debatac_atac_peak_annotation_df}
# peak annotation
atac.peak.anno <- AnnoPeak(
  peak.df = atac.peak.df, species = "Mouse",
  seq.style = "UCSC", up.dist = 20000, down.dist = 20000
)
atac.peak.anno.df = atac.peak.anno$df
head(atac.peak.anno.df)
```

```{r debatac_atac_peak_annotation_plot, fig.height = 14, fig.width = 14, fig.align = "center", warning=FALSE}
atac.peak.anno$plots
```

<hr />

## Integrate ATAC-seq and RNA-seq
In this step, we will integrate ATAC-seq and RNA-seq to explore the role of chromatin structure of promoter region in regulating target gene expression.
### Integrate
```{r debatac_integrate}
debatac.res = DEbPeak(de.res = debatac.dds.results.ordered, peak.res = atac.peak.anno.df, peak.anno.key = "Promoter", merge.key="SYMBOL")
head(debatac.res)
```

<hr />

### Integrate summary
```{r debatac_integrate_plot, fig.height = 4, fig.width = 4, fig.align = "center", warning=FALSE}
# DE and ATAC venn plot
debatac.plot = PlotDEbPeak(debatac.res, peak.type = "ChIP", show_percentage=FALSE)
debatac.plot
```

<hr />

## Functional enrichment
```{r debatac_integrate_fe}
# functional enrichment on direct targets
debatac.fe.results = DEbPeakFE(de.peak = debatac.res, peak.type = "ATAC", gene.type = "ENTREZID", species="Mouse",save = F)
```

### FE on up-regulated targets 
```{r debatac_integrate_fe_up, warning=FALSE, fig.height = 20, fig.width = 8, fig.align = "center"}
up.debatac.fe.results=debatac.fe.results[["UPbATAC"]][["GO"]]
head(up.debatac.fe.results[["table"]])

up.debatac.fe.results[["plot"]]
```

<hr />

### FE on down-regulated targets
```{r debatac_integrate_fe_down, warning=FALSE, fig.height = 20, fig.width = 8, fig.align = "center"}
down.debatac.fe.results=debatac.fe.results[["DOWNbATAC"]][["GO"]]
head(down.debatac.fe.results[["table"]])

down.debatac.fe.results[["plot"]]
```

<hr />

## Find motif
[Different from ChIP-seq which profile specific protein–DNA interactions (depends on the antibody provided), ATAC-seq can proflie global chromatin accessibility](https://link.springer.com/article/10.1186/s43556-020-00009-w). Thus, the peaks of ATAC-seq may originate from hundreds of TFs. To reveal the possible TFs involved, motif enrichment analysis is a good solution.
```{r debatac_integrate_motif, eval = FALSE}
# debatac.res: similar to debatac.res
# this step is time consuming!
ATAC.motif = FindMotif(inte.res = debatac.res, peak.anno.res = peak.anno.df, gene.key = "SYMBOL",
                       homer.motif.path = '~/anaconda3/bin/findMotifsGenome.pl',
                       genome = '/path/to/genome.fa',
                       out.folder = NULL,other.paras = '-len 8,10,12 -size -100,50 -S 25')
```

<hr />

## Save results
Save results for downstream analysis:
```{r save_image}
save.image(file = "/home/songyabing/R/learn/tmp/DEbPeak/RNAandATAC.RData")
```

<hr />

## Session info
```{r session}
sessionInfo()
```

<hr />


