---
title: "IntegrateChIP"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Gene expression is controlled by transcription factors (TFs), which [binding to DNA recognition sites located near their target genes](https://www.nature.com/articles/s41467-019-14217-8). [Chromatin Immunoprecipitation Sequencing (ChIP-Seq) can be used to profile genome-wide protein–DNA interactions](https://www.nature.com/articles/nrg2641), such as identifying sites of TFs binding. Therefore, ChIP-Seq can be used to [reveal mechanisms involved in differential gene regulation](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4614920/).

`DEbPeak` implements functions to integrate RNA-seq and ChIP-seq, which can be used to investigate the direct and indirect targets of a TF, thus revealing the regulation network of gene expression.

<hr />

## Example data
The data used here contains RNA-seq and ChIP-seq datasets from [RUNX represses Pmp22 to drive neurofibromagenesis](https://www.science.org/doi/10.1126/sciadv.aau8389): 

* RNA-seq: two genotypes and three samples per genotype, the raw data are stored in [GSE122774](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122774)
* ChIP-seq: two genotypes and one sample per genotype, the raw data are stored in [GSE122775](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122775)

## Differential Expression Analysis (RNA-seq)
```{r debchip_degs}
# library
suppressWarnings(suppressMessages(library(DESeq2)))
suppressWarnings(suppressMessages(library(DEbPeak)))

# prepare count matrix and metadata
debchip.count.file <- system.file("extdata", "debchip_count.txt", package = "DEbPeak")
debchip.meta.file <- system.file("extdata", "debchip_meta.txt", package = "DEbPeak")
debchip.count.matrix <- read.table(file = debchip.count.file, header = T, sep = "\t")
debchip.meta.info <- read.table(file = debchip.meta.file, header = T)
# create DESeqDataSet object
debchip.dds <- DESeq2::DESeqDataSetFromMatrix(countData = debchip.count.matrix, 
                                              colData = debchip.meta.info, 
                                              design = ~condition)

# set control level
debchip.dds$condition <- relevel(debchip.dds$condition, ref = "NF")
# conduct differential expressed genes analysis
debchip.dds <- DESeq(debchip.dds)
# extract results
debchip.dds.results <- results(debchip.dds,contrast=c("condition",'RX','NF'))
debchip.dds.results.ordered <- debchip.dds.results[order(debchip.dds.results$log2FoldChange,decreasing = TRUE),]
head(debchip.dds.results.ordered)
```

<hr />

## Process ChIP-seq data
### Consensus peaks
Due to the high noise of ChIP-seq experiments, control samples can be used to remove possible nonspecific interactions. Besides control samples, [replicates are another way to separate actual biological events from variability resulting from random chance](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3962196/). The peaks which have high consistency between replicates are likely to be genuine signals, and we call these peaks consensus peaks. In this step, we will get consensus peaks with [MSPC](https://genometric.github.io/MSPC/docs/quick_start/) when multiple peak files are available, but when there is only one peak file, we will use it directly (**make sure this peak file contains five columns: "chr", "start", "stop", "name", "score"**).
```{r debchip_consensus_peak}
# get consensus peak
peak.file = system.file("extdata", "debchip_peaks.bed", package = "DEbPeak")
peak.df = GetConsensusPeak(peak.file = peak.file)
head(peak.df)
```

<hr />

### Peak profile
Check the profle of consensus peaks:
```{r debchip_peak_profile, fig.height = 6, fig.width = 10, fig.align = "center", warning=FALSE}
# peak profile plot
peak.profile = PeakProfile(peak.df ,species="Mouse", by = "gene", region.type = "body", nbin = 800)
peak.profile$profile.plot
```

<hr />

### Peak annotation
In this step, we will perform peak annotation with [ChIPseeker](https://bioconductor.org/packages/release/bioc/vignettes/ChIPseeker/inst/doc/ChIPseeker.html). This will include annotating peaks with its nearby genes, assigning genomic region of the peaks, et al.
```{r debchip_peak_annotation_df}
# peak annotation
peak.anno = AnnoPeak(peak.df = peak.df,species = "Mouse",seq.style = "UCSC",up.dist = 20000,down.dist = 20000)
peak.anno.df = peak.anno$df
head(peak.anno.df)
```

```{r debchip_peak_annotation_plot, fig.height = 14, fig.width = 14, fig.align = "center", warning=FALSE}
peak.anno$plots
```

The above pie plot is not based on `ggplot2`, here we provide function to create pie plot based on `ggplot2`:
```{r debchip_peak_annotation_pie, fig.height = 10, fig.width = 10, fig.align = "center", warning=FALSE}
PeakAnnoPie(peak.anno$anno.obj)
```

<hr />

## Integrate ChIP-seq and RNA-seq
In this step, we will integrate ChIP-seq and RNA-seq to get plausible direct targets of TF (in this example, Runx).
### Integrate
```{r debchip_integrate}
debchip.res = DEbPeak(de.res = debchip.dds.results.ordered, peak.res = peak.anno.df, peak.anno.key = "Promoter", merge.key="SYMBOL")
head(debchip.res)
```

<hr />

### Integrate summary
```{r debchip_integrate_plot, fig.height = 4, fig.width = 4, fig.align = "center", warning=FALSE}
# DE and ChIP venn plot
# debchip.plot = PlotDEbPeak(debchip.res, peak.type = "ChIP", show_percentage=FALSE)
# debchip.plot
debchip.plot = InteVenn(inte.res = debchip.res, inte.type = "DEbPeak", peak.type = "ChIP", 
                        peak.mode = "consensus", gene.col = "SYMBOL",show_percentage = FALSE)
debchip.plot
```

<hr />

## Functional enrichment
There are **five categories** for users to choose to perform functional enrichment: `UP`, `Peak`, `DOWN`, `DOWNbPeak`, `UPbPeak`. Here, we will use `DOWNbPeak` and `UPbPeak` as examples.

### UPbPeak
```{r debchip_integrate_fe_up}
# functional enrichment on direct targets
# debchip.up.fe.results = DEbPeakFE(de.peak = debchip.res, peak.fe.key = "UPbPeak", 
#                                   gene.type = "ENTREZID", species="Mouse",save = F)
debchip.up.fe.results = InteFE(inte.res = debchip.res, fe.key = "UPbPeak", inte.type = "DEbPeak",
                               gene.type = "ENTREZID", species="Mouse",save = F)
```

The results:
```{r debchip_integrate_fe_up_show, warning=FALSE, fig.height = 20, fig.width = 8, fig.align = "center"}
# the result table
head(debchip.up.fe.results[["GO"]][["table"]])
# the result plot
debchip.up.fe.results[["GO"]][["plot"]]
```

<hr />

### DOWNbPeak
```{r debchip_integrate_fe_down}
# functional enrichment on direct targets
# debchip.down.fe.results = DEbPeakFE(de.peak = debchip.res, peak.fe.key = "DOWNbPeak", 
#                                     gene.type = "ENTREZID", species="Mouse",save = F)
debchip.down.fe.results = InteFE(inte.res = debchip.res, fe.key = "DOWNbPeak", inte.type = "DEbPeak",
                                 gene.type = "ENTREZID", species="Mouse",save = F)
```

The results:
```{r debchip_integrate_fe_down_show, warning=FALSE, fig.height = 20, fig.width = 8, fig.align = "center"}
# the result table
head(debchip.down.fe.results[["GO"]][["table"]])
# the result plot
debchip.down.fe.results[["GO"]][["plot"]]
```

<hr />

## Save results
Save results for downstream analysis:
```{r save_image}
save.image(file = "/home/songyabing/R/learn/tmp/DEbPeak/RNAandChIP.RData")
```

<hr />

## Session info
```{r session}
sessionInfo()
```

<hr />



